Results section

### 2022-10-19 code/Matrinem_data_analysis.Rmd

---
title: "Results section"
author: "Kaare"
date: "18/10/2022"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r loading libraries}
library(readxl)
library(tidyverse)
library(broom)
library(ggpubr)
#library(AICcmodavg)
library(visdat)
library(rstatix)
library(gt)
library(gtsummary)
library(patchwork)
library(jtools)
```

@@

## Preface

## Introduction

This thesis is about...

## Background

## Methods and Materials

Prior to results are presented a short description of the studies conducted needs to be made.

@@

+-------------+-------------------------------------------------------------------------------+----------------------------------------------------------------+------------------------------------------------------------------------+-----------------------------------------------------------------+
|             | PreExp 1                                                                      | PreExp 2                                                       | PreExp 3                                                               | NEC-FvT                                                         |
+=============+===============================================================================+================================================================+========================================================================+=================================================================+
| **Aim**     | To test feasibility of formula feeding with PICC line                         | Selectively targeting gram positive vs. gram negative bacteria | Testing if there is an interaction-effect with feeding and antibiotics | Combining reults from PreExp 1-3 into final experimental setup. |
+-------------+-------------------------------------------------------------------------------+----------------------------------------------------------------+------------------------------------------------------------------------+-----------------------------------------------------------------+
| Protocols   | Testing a broad spectrum antibiotics mix in combination with formula feeding. | Groups                                                         | Groups                                                                 | Groups                                                          |
|             |                                                                               |                                                                |                                                                        |                                                                 |
|             | 4 groups                                                                      |                                                                |                                                                        |                                                                 |
|             |                                                                               |                                                                |                                                                        |                                                                 |
|             | 1.  CON-BF                                                                    |                                                                |                                                                        |                                                                 |
|             | 2.  CON-FF                                                                    |                                                                |                                                                        |                                                                 |
|             | 3.  AB-BF                                                                     |                                                                |                                                                        |                                                                 |
|             | 4.  AB-FF                                                                     |                                                                |                                                                        |                                                                 |
+-------------+-------------------------------------------------------------------------------+----------------------------------------------------------------+------------------------------------------------------------------------+-----------------------------------------------------------------+
| N           | 38 (32 survived)                                                              | 18                                                             | 32                                                                     | 26                                                              |
+-------------+-------------------------------------------------------------------------------+----------------------------------------------------------------+------------------------------------------------------------------------+-----------------------------------------------------------------+

@@

@@

-   Can we keep offspring alive with formula feeding?
-   Will combining MAT with FF result in a more inflammatory intestine as measured by cytokine analysis in the ileum.

Experiment 1 was a pilot to test whether it was possible to feed 3-day old mice with the PICC. The experimental setup followed the standard setup. Antibiotics was administered according to the description in Chen et al., though with a lower dose based on previous experiments. Our primary study objective was survival. Secondary objective was macroscopic evaluation of intestines for inflammation, intestinal distention and similar NEC-like symptoms. One preliminary hypothesis was a reduction in weight of offspring from antibiotics-treated mothers.

**Results**:

```{r include=FALSE}
preexp_1 <- read_excel("../data/processed/weight_development.xlsx") %>% 
  filter(exp_number == "1")

```{r}
#__________Load in cytokine data for all experiments________
cytokines_1 <- read_excel(path = "../data/processed/cytokines.xlsx") %>%  
  filter(exp_number == 1) %>% 
  filter(tissue == "ileum") %>% 
  mutate(
    group = recode_factor(  #Change the order of the groups to desired plot order
      group,
      "CON-BF"= "CON-BF",
      "CON-FF" = "CON-FF",
      "AB-BF"= "AB-BF",
      "AB-FF"= "AB-FF"
      ))
#___________________________________________________________

@@

`r nrow(preexp_1)` mice were born to 6 different mothers. Of these 32 survived and were included in the analysis.

`r nrow(cytokines_1)` samples were available for analysis of concentrations of the proinflammatory cytokines `r names_of_cytokines`. Concentrations were measured in 1-cm sections of proximal/distal ileum tissue using MSD Mesoscale Kits. Distributions were assessed for normality with Shapiro-Wilk Normality Test (see table below), and plot of data were made to assess distributions visually:

```{r echo=FALSE, fig.cap='Table:}
normality_result_1
```

@@

  tbl1_gt <- tbl1 %>% 
  as_gt() %>%
  tab_header(title = md("**Proinflammatory cytokines**")) %>%
  opt_align_table_header("center") %>% 
  tab_options(heading.title.font.size = 20,
              heading.background.color = "lightgrey",
              ) %>% 
  tab_source_note(
    source_note = " * P-value indicates overall group differences"
  ) %>%
  tab_source_note(
    source_note = md("**Matrinem round 1**")
  ) %>%
  cols_width(label ~ px(100),
             stat_1 ~ px(150),
             stat_2 ~ px(150),
             stat_3 ~ px(150),
             stat_4 ~ px(150))
# Save and export files
# tbl1 %>%  #Save as excel file.
#   as_hux_table() %>%
#   huxtable::quick_xlsx(., file = "../Results/Round 1/ileum-cytokines.xlsx")
# 
# tbl1 %>% #Save as image file.
# as_flex_table() %>%
#   flextable::save_as_image(path = "../../Results/Round 1/ileum_cytokines_table.png",webshot = "webshot")

@@

The median concentration of the various cytokines are shown in the table below:

```{r, echo=FALSE}
tbl1_gt
```

@@

@@

@@

```{r echo=FALSE, warning=FALSE}
reg_mod_cyt_1 %>% 
  select(-statistic) %>% 
  knitr::kable(., caption = 'Significant effectors on cytokine expression', digits = 3)
```

**The conclusion** is that many of the cytokines show an interaction effect of the two treatment parameters

@@

-   Can we keep offspring alive with formula feeding?

    -   After fine-tuning the feeding method we we able to successfully feed all surviving (32) mice every 3 hours.
    -   The formulafed animals gained weight - albeit less than their breastfed peers - during the 48 hours.
    -   MAT did not affect growth. CON-BF and AB-BF gained weight at a similar pace.

@@

-   Will combining MAT with FF result in a more inflammatory intestine as measured by cytokine analysis in the ileum.

>>>>

---
title: "Results section"
author: "Kaare"
date: "6/16/2022"
output:
  pdf_document: default
  html_document: default
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
#setwd(dir = "../../2022_matrinem_thesis/")
```

```{r loading libraries, message=FALSE}
library(readxl)
library(tidyverse)
library(broom)
library(ggpubr)
#library(AICcmodavg)
library(visdat)
library(rstatix)
library(gt)
library(gtsummary)
library(patchwork)
library(jtools)
```

@@

## The research questions

@@

+-----------+-------------------------------------------------------------------------------+----------+----------+---------+
|           | PreExp 1                                                                      | PreExp 2 | PreExp 3 | NEC-FvT |
+===========+===============================================================================+==========+==========+=========+
| **Aim**   | To test feasibility of formula feeding with PICC line                         |          |          |         |
+-----------+-------------------------------------------------------------------------------+----------+----------+---------+
| Protocols | Testing a broad spectrum antibiotics mix in combination with formula feeding. |          |          |         |
|           |                                                                               |          |          |         |
|           | 4 groups                                                                      |          |          |         |
|           |                                                                               |          |          |         |
|           | 1.  CON-BF                                                                    |          |          |         |
|           | 2.  CON-FF                                                                    |          |          |         |
|           | 3.  AB-BF                                                                     |          |          |         |
|           | 4.  AB-FF                                                                     |          |          |         |
+-----------+-------------------------------------------------------------------------------+----------+----------+---------+
| N         | 38 (32 survived)                                                              |          |          |         |
+-----------+-------------------------------------------------------------------------------+----------+----------+---------+

@@

#Load in the data

```{r include=FALSE}
preexp_1 <- read_excel("data/processed/weight_development.xlsx") %>% 
  filter(exp_number == "1") 
```

```{r}
#__________Load in cytokine data for all experiments________
cytokines_1 <- read_excel(path = "data/processed/cytokines.xlsx") %>%  
  filter(exp_number == 1) %>% 
  filter(tissue == "ileum") %>% 
  mutate(
    group = recode_factor(  #Change the order of the groups to desired plot order
      group,
      "CON-BF"= "CON-BF",
      "CON-FF" = "CON-FF",
      "AB-BF"= "AB-BF",
      "AB-FF"= "AB-FF"
      ))
#___________________________________________________________

@@

##### Can we keep offspring alive with formula feeding?

`r nrow(preexp_1)` mice were born to 6 different mothers. Of these 32 survived and were included in the analysis. After fine-tuning the feeding method we we able to successfully feed all surviving (32) mice every 3 hours for 2 days (48 hours).

##### Will MAT and FF affect levels of inflammatory cytokines in ileum tissue?

@@

`r nrow(cytokines_1)` samples were available for analysis of concentrations of the proinflammatory cytokines `r names_of_cytokines`. Concentrations were measured in 1-cm sections of proximal/distal ileum tissue using MSD Mesoscale Kits.

@@

  tbl1_gt <- tbl1 %>% 
  as_gt() %>%
  tab_header(title = md("**Proinflammatory cytokines**")) %>%
  opt_align_table_header("center") %>% 
  tab_options(heading.title.font.size = 20,
              heading.background.color = "lightgrey",
              ) %>% 
  tab_source_note(
    source_note = " * P-value indicates overall group differences"
  ) %>%
  tab_source_note(
    source_note = md("**Matrinem round 1**")
  ) %>%
  cols_width(label ~ px(100),
             stat_1 ~ px(150),
             stat_2 ~ px(150),
             stat_3 ~ px(150),
             stat_4 ~ px(150))
```

The concentrations of the various cytokines are shown in the table below:

```{r, echo=FALSE}
tbl1_gt

@@

@@

Cytokine levels are shown below:

```{r echo=FALSE}
all_cytokines_patch
```

Distributions were assessed for normality with Shapiro-Wilk Normality Test (see table below), and plot of data were made to assess distributions visually:

```{r echo=FALSE, fig.cap='Table of test results:'}
normality_result_1
```

Most measured cytokines wore not normally disrtibuted. Consider logtransformation

@@

The plot above shows that the cytokines significantly affected / difference were `r tbl1_p$Cytokines`.

@@

```{r echo=FALSE, warning=FALSE}
reg_mod_cyt_1 %>% 
  select(-statistic) %>% 
  knitr::kable(., caption = 'Significant effectors on cytokine expression', digits = 5)

ff_affected_cytokines <- reg_mod_cyt_1 %>% 
  filter(term %in% "type_of_feedFF") %>% 
  pull(cytokines)

ff_affected_cytokines_pvals <- reg_mod_cyt_1 %>% 
  filter(term %in% "type_of_feedFF") %>% 
  pull(p.value)
```

Formula-feeding looks to be the most determining factor for changes to cytokine expression.

**The conclusion** is that many of the cytokines show an interaction effect of the two treatment parameters.

@@

-   We could keep them alive with proper feeding technique.

-   Formula feeding looks to be the most determining factor for `r ff_affected_cytokines` (p-values `r ff_affected_cytokines_pvals`, respectively).

@@

#### **Rationale** 
- Based on the results from Preexp 1 we knew that we could perform the experimental setup without too many casualties. 
- We had seen that formula-feeding did affect the expression of cytokines, but visual inspection of the intestines showed that we had not managed to create a NEC-like phenotype. 
- After considering the results we realized that our assumptions were based on a dysbiotics microbiome in the mother leading to less immunizing milk. However, we had likely not created a dysbiosis as much as an overall depletion by administering a broad-spectered antibiotic. 
We therefore hypothezised that targeting specific stains of bacteria would lead to a more dysbiotic microbiome and allow for greater disbiosis in the offspring.

- In **PreExp 2** we therefore tested whether selectively targeting gram+ or gram- bacteria would lead to different and more pronounced NEC-like symptoms.

#### Research questions: 
##### Will selectively targering gram+ or gram- bacteria with Vancomycin and Gentamicin respectively lead to more pronounced NEC-like symptoms compared with no treatment on the background of breastfeeding.

### Experiment 4 (NEC-FvT)

The PreExps are now over and it is time to conduct the study. Having determined the optimal experimental protocol in PreExp 1-3 we are ready with the design.

#### Overview - Experimental Setup

#### Intestinal permeability - FITC

```{r FITC load data, include=FALSE}
fitc <- read_excel("data/processed/fitc.xlsx") %>% 
  rename(values = `result_µg/ml_undiluted`) %>% 
  drop_na(values)
```

Intestinal permeability is assessed with the FITC-Dextran assay. First I inspect the distribution of results.

```{r FITC check for normality}
fitc %>% 
  filter(exp_number == 4) %>% 
  group_by(group) %>% 
  shapiro_test(values)
```

As results and highly skewed I try transforming the FITC results to log10.

```{r FITC try log transforming data}
fitc <- fitc %>% 
  mutate(log_10_values = log10(values))  # Logtransform results

```{r FITC check for normality log transform}
fitc %>% 
  filter(exp_number == 4) %>% 
  group_by(group) %>% 
  shapiro_test(log_10_values)
```

```{r FITC visual inspections of data distribution}
#Density plots non-log
density <- fitc %>% 
  filter(exp_number == 4) %>% 
  ggplot(aes(x = values, fill = group)) + 
  geom_density() +
  facet_wrap(~ group,
             scales = "free")

#Jitter plots non-log
jitter <- fitc %>% 
  filter(exp_number == 4) %>% 
  ggplot(aes(x = group, y = values)) + 
  geom_jitter() +
  facet_wrap(~ group,
             scales = "free")

#Density plots log
density <- fitc %>% 
  filter(exp_number == 4) %>% 
  ggplot(aes(x = , fill = group)) + 
  geom_density() +
  facet_wrap(~ group,
             scales = "free")

#Jitter plots non-log
jitter <- fitc %>% 
  filter(exp_number == 4) %>% 
  ggplot(aes(x = group, y = values)) + 
  geom_jitter() +
  facet_wrap(~ group,
             scales = "free")
